---
title: 'L06: Intro to Linear Regression:  code from lecture '
output: pdf_document
---

<!-- libraries -->
```{r,include=FALSE,purl=FALSE}
library(knitr) # for include_graphics() 
library(dplyr)
```

## Fitting a linear model in R

Code template:

```{r lm-template, eval=F}
lm(formula = y ~ x, data = your_dataset)
```

- `lm()` is the function for a linear model.
- The first argument that `lm()` wants is a formula `y ~ x`.
    - `y` is the response variable from your dataset
    - `x` is the explanatory variable
    - be careful with the order of x and y! It is opposite from the default order in ggplot 
    
    ggplot(data,aes(x=your_x, y=your_y))
    
- The second argument sent to `lm()` is the data set.
    - the default order of declaring the data as the second argument in lm() is different from the ggplot2 and dplyr functions



### Why the package broom?

We will pull in a new package here:
library(broom) 
and apply the tidy() function as follows:
tidy(your_lm) 

* `broom` has functions that make the output from the linear model look clean 
* `tidy` is a function from the `broom` package that tidies up the output


### Example: Manatee deaths and powerboat purchases 

Let's apply the `lm()` function. Recall the manatee example from our last lecture that 
examined the relationship between the number of registered `powerboats` and the 
number of manatee `deaths` in Florida between 1977 and 2016.

```{r load-mana-data, echo=F, message=FALSE, warning=FALSE}
library(readr)
mana_data <- read_csv("Ch03_Manatee-deaths.csv")
```

Recall that the relationship appeared linear when we examined the scatter plot:

```{r scatter-plot-mana, out.width="80%", message=FALSE}
library(ggplot2)
mana_death1<-ggplot(mana_data, aes(x = powerboats, y = deaths)) +
  geom_point() +
  theme_minimal(base_size = 15)
```

### Manatee deaths and powerboat purchases 
```{r mana_plot1, out.width="80%"}
mana_death1
```

### lm() of manatee deaths and powerboat purchases
Calculate the line of best fit: 
```{r run-linear-model-mana, out.width="80%"}
mana_lm <- lm(deaths ~ powerboats, mana_data)
library(broom) 
tidy(mana_lm) 
```

Only pay attention to the term and estimate columns for now.

### lm() of manatee deaths and powerboat purchases

**Interpret the model output**
```{r manout, out.width="80%", echo=F}
tidy(mana_lm)
```
- Intercept: The predicted number of deaths if there were no powerboats. But the
prediction is negative. Why?

- Powerboats: This is the slope. What does the estimated slope for powerboats mean?

### Add the regression line to the scatter plot using `geom_abline()`
We add a statement to our ggplot
  geom_abline(intercept = your_intercept, slope = your_slope)
  
  so for our manatee data
  geom_abline(intercept = -46.7520, slope = 0.1358)

Note:  by default, ggplot only shows the ploting region that corresponds to the range of data

### Add the regression line to the scatter plot using `geom_abline()`

```{r scatter-with-line-of-best-fit, echo=FALSE}
  #students, know how to use geom_abline() to add a line to scatterplot
  #students do not need to know the coord_fixed command.
default_zoom_plot <- ggplot(mana_data, aes(x = powerboats, y = deaths)) +
  geom_point() + 
  labs(x = "Powerboats registered (X 1000)",
       y = "Manatee deaths") +
  geom_abline(intercept = -46.7520, slope = 0.1358) + 
  coord_fixed(ratio = 5) +
  theme_minimal(base_size = 15)
```

```{r zoom-out-plot, out.width="90%", echo = F}
zoom_out_plot <- ggplot(mana_data, aes(x = powerboats, y = deaths)) +
  geom_hline(yintercept = 0, col = "forest green") + 
  geom_vline(xintercept = 0, col = "forest green") +
  geom_point() + 
  labs(x = "Powerboats registered (X 1000)",
       y = "Manatee deaths") +
  geom_abline(intercept = -46.7520, slope = 0.1358) + 
  scale_x_continuous(limits = c(0, 1050)) + 
  scale_y_continuous(limits = c(-50, 110)) + 
  coord_fixed(ratio = 6) +
  theme_minimal(base_size = 15)
  #students, know how to use geom_abline() to add a line to scatterplot
  #students do not need to know the coord_fixed command.

library(patchwork) #don't need to know this package
both_plots <- default_zoom_plot + zoom_out_plot + plot_layout() #don't need to know this code
both_plots
```


### Interpreting the slope

```{r, out.width="90%", echo=F}
tidy(mana_lm)
```

- Powerboats: This is the slope. What does the estimated slope for powerboats mean?

- A one unit change in the number of powerboats registered (X 1,000) is 
associated with an increase of manatee deaths of 0.1358. That is, an increase in 
the number of powerboats registered by 1,000 is association with 0.1358 more 
manatee deaths.
- If powerboat registered increased by 100,000 how many more manatee deaths are
expected?

### Change units
```{r, out.width="80%", message=F}
mana_data_units<-mana_data%>%mutate(actual_powerboats = powerboats * 1000)
mana_lm_units <- lm(deaths ~ actual_powerboats, mana_data_units)
tidy(mana_lm_units)
```

What happened to the slope?  To the intercept?

### Getting the R-squared from your model

When we run a linear model, the r-squared is also calculated. Here is how to see the r-squared for the manatee data:

```{r, glance-at-model}
library(broom)
glance(mana_lm)
```

Focus on:

- Column called `r.squared` values only.
- Interpretation of r-squared: The fraction of the variation in the values of y
that is explained by the line of best fit.


### Correlation vs R Squared
```{r calc-corr, warning=FALSE, message=FALSE}
library(dplyr)
mana_cor <- mana_data %>% 
  summarize(corr_mana = cor(powerboats, deaths))
mana_cor
```

### Correlation vs R Squared
```{r calc-corr2, warning=FALSE, message=FALSE}
glance(mana_lm)%>% pull(r.squared)
#square the correlation coefficient
.9448054^2
```




## Transforming data

### Transforming data
- Sometimes, the data is transformed to another scale so that the relationship
between the transformed $x$ and $y$ is linear
- Table 3.4 in B&M provides data on the mean number of seeds produced in a year 
by several common tree species and the mean weight (in milligrams) of the seeds 
produced. 

```{r make-seed-dataset, echo=FALSE, warning=FALSE}
library(tibble)
seed_data <- tribble(~ species, ~ seed_count, ~ seed_weight,
                       "Paper birch", 27239, 0.6,
                       "Yellow birch", 12158, 1.6,
                       "White spruce", 7202, 2.0,
                       "Engelman spruce", 3671, 3.3, 
                       "Red spruce", 5051, 3.4, 
                       "Tulip tree", 13509, 9.1, 
                       "Ponderosa pine", 2667, 37.7, 
                       "White fir", 5196, 40.0, 
                       "Sugar maple", 1751, 48.0, 
                       "Sugar pine", 1159, 216.0, 
                       "American beech", 463, 247, 
                       "American beech", 1892, 247,
                       "Black oak", 93, 1851, 
                       "Scarlet oak", 525, 1930, 
                       "Red oak", 411, 2475, 
                       "Red oak", 253, 2475,
                       "Pignut hickory", 40, 3423, 
                       "White oak", 184, 3669, 
                       "Chestnut oak", 107, 4535)

```

### Scatter plot of `seed_weight` vs. `seed_count`
```{r, out.width="70%", echo=FALSE}
ggplot(seed_data, aes(seed_count, seed_weight)) + 
  geom_point() +
  theme_minimal(base_size = 15)
```

- `seed_count` and `seed_weight` both vary widely
- Their relationship is not linear

### Investigate the relationship between their logged variables

- Add transformed variables to the dataset using `mutate()`. 
- We add both log base $e$ and log base 10 variables for illustration

```{r calc-logged-vars, message=F, warning=F, out.width="80%"}
library(dplyr)
seed_data <- seed_data %>% mutate(log_seed_count = log(seed_count), 
                                  log_seed_weight = log(seed_weight),
                                  log_b10_count = log(seed_count, 10),
                                  log_b10_weight = log(seed_weight, 10)) 
```

### Plot transformed data (log base e)

```{r scatter-logged, out.width="50%", echo=FALSE}
ggplot(seed_data, aes(log_seed_count, log_seed_weight)) + 
  geom_point() +
  labs(x = "Log of seed count", y = "Log of seed weight", 
       title = "Using the natural log (base e)") +
  theme_minimal(base_size = 15)
```

### Plot transformed data (log base 10)

```{r scatter-log-base10, out.width="50%", echo=FALSE}
ggplot(seed_data, aes(log_b10_count, log_b10_weight)) + 
  geom_point() +
  labs(x = "Log of seed count", y = "Log of seed weight", 
       title = "Using log base 10") +
  theme_minimal(base_size = 15)
```

- You can use either base 10 or base $e$ for class.
- The calculations using base $e$ are easier
- If you are calculating in a different program or on a calculator make sure you know what base is being used!

### `lm()` on the log (base e) variables

```{r run-linear-model-seed}
seed_mod <- lm(log_seed_weight ~ log_seed_count, data = seed_data)
tidy(seed_mod)
glance(seed_mod) %>% pull(r.squared)
```

- Interpret the intercept:
- Interpret the slope: 

### `lm()` on the log (base 10) variables

```{r run-linear-model-seed-b10}
seed_mod_b10 <- lm(log_b10_weight ~ log_b10_count, data = seed_data)
tidy(seed_mod_b10)
glance(seed_mod_b10) %>% pull(r.squared)
```

- What is different from the log base $e$ output?


## How do outliers affect the line of best fit?

### How do outliers affect the line of best fit?

To study this, we use data from the Organization for Economic Co-operation and 
Development (OECD). This dataset was downloaded from http://dx.doi.org/10.1787/888932526084 
and contains information on the health expenditure per capita and the GDP per
capita for 40 countries. 

```{r read-spending-data, warning=FALSE}
library(readxl)

spending_dat <- read_xlsx("Ch04_Country-healthcare-spending.xlsx", 
                          sheet = 2,
                          range = "A7:D47")
```

### Have a look

Next, we want to examine the imported data to see if it is how we expect:
```{r examine-spending-data, out.width="80%"}
head(spending_dat)
```

### Rename() some variables to use a consistent naming style

If the variable name has spaces, we must use back ticks when referring to it:

```{r rename-variables-with-spaces}
library(dplyr)
spending_dat <- spending_dat %>% 
  rename(country_code = Country.code,
         health_expenditure = `Health expenditure per capita`, # back ticks
         GDP = `GDP per capita`) # back ticks
```

### Examine the relationship

Make a scatter plot of `health_expenditure` (our response variable) vs. each country's level of `GDP`:

```{r plot-spending-data, message=FALSE, echo=FALSE, warning=FALSE, out.width="80%"}
library(ggrepel) 
#this library is used for adding labels to a scatter plot that don't overlap the data points
ggplot(spending_dat, aes(x = GDP, y = health_expenditure)) + 
  geom_point() +
  geom_text_repel(aes(label = country_code)) +
  theme_minimal(base_size = 15)
```

### Examine the relationship

Is the relationship linear? Which countries are outliers? 

Fit a linear model to these data 

```{r fit-lm-and-plot}
lm(health_expenditure ~ GDP, data = spending_dat)
```

### Examine the relationship
Add the regression line to the graph:
```{r plotgdp, out.width="80%"}
GDP_withline<-ggplot(spending_dat, aes(x = GDP, y = health_expenditure)) + 
  geom_point() +
  geom_text_repel(aes(label = country_code)) + # this adds the country code as a label 
  geom_abline(intercept = 44.65623, slope = 0.09399, lty = 2) +
  theme_minimal(base_size = 15)
```

### Examine the relationship
```{r showplotgdp, out.width="80%", echo=F}
GDP_withline
```

### Examine the relationship without Luxembourg in the data

Let's see whether removing Luxembourg changes the fit of the line. We can remove 
Luxembourg using the `filter()` command from `dplyr`:

```{r remove-lux-and-save-new-dataset}
spending_dat_no_LUX <- spending_dat %>% filter(country_code != "LUX")

lm(health_expenditure ~ GDP, data = spending_dat_no_LUX)
```


### Examine the relationship without Luxembourg in the data
```{r noluxplot,out.width="80%"}
GDP_nolux<-ggplot(spending_dat, aes(x = GDP, y = health_expenditure)) + geom_point() +
  geom_text_repel(aes(label = country_code)) + 
  geom_abline(intercept = 44.65623, slope = 0.09399, lty = 2) + 
  geom_abline(intercept = -785.1044, slope = 0.1264, col = "red") +
  theme_minimal(base_size = 15)
```

### Examine the relationship without Luxembourg in the data
```{r showgdpnolux, echo=F, out.width="80%"}
GDP_nolux
```

### Examine the relationship without USA in the data

```{r remove-usa-and-save-new-dataset}
spending_dat_no_USA <- spending_dat %>% filter(country_code != "USA")

lm(health_expenditure ~ GDP, data = spending_dat_no_USA)
```


### Examine the relationship without USA in the data
```{r nousnoluxplot, out.width="80%"}
GDP_nousa<-ggplot(spending_dat, aes(x = GDP, y = health_expenditure)) + geom_point() +
  geom_text_repel(aes(label = country_code)) + 
  geom_abline(intercept = 44.65623, slope = 0.09399, lty = 2) + 
  geom_abline(intercept = 152.26274, slope = 0.08714, col = "blue") +
  theme_minimal(base_size = 15)
```

### Examine the relationship without USA in the data
```{r shownousplot, out.width="80%", echo=F}
GDP_nousa
```

### Examine the relationship without LUX or USA in the data

Let's write the code together to remove both the USA and LUX and see how it 
affects the fit:

```{r remove-usa-lux}
spending_dat_no_USA_LUX <- spending_dat %>% 
  
  filter(country_code != "USA" & country_code != "LUX")

#alternatively, you could have written:
spending_dat_no_USA_LUX <- spending_dat %>%
  
  filter(! country_code %in% c("USA", "LUX"))

#pick the filter command that makes the most sense to you.
```

### Examine the relationship without LUX or USA in the data

```{r modelnoluxnouse}
lm(health_expenditure ~ GDP, data = spending_dat_no_USA_LUX)
GDP_noluxnousa<-ggplot(spending_dat_no_USA_LUX, aes(x = GDP, y = health_expenditure)) + geom_point() +
  geom_text_repel(aes(label = country_code)) + 
  geom_abline(intercept = 44.65623, slope = 0.09399, lty = 2) + 
  geom_abline(intercept = -592.6973, slope = 0.1166 , col = "green") +
  theme_minimal(base_size = 15)
```


### Examine the relationship without LUX or USA in the data
```{r shownoluxnouse, out.width="80%", echo=F}
GDP_noluxnousa
```



## Counfounding 

### Discussion of some examples from Baldi & Moore
We can encode these differences in a causal diagram. Here is a simple one to demonstrate the concept:
```{r dag, echo=F, fig.height=1.5, fig.width=3, warning=FALSE}
#students, you don't need to know the dagitty package or understand this code
library(dagitty)

g <- dagitty('dag {
    Music.Lessons [pos="0,1"]
    High.grades [pos="2,1"]
    Family.income [pos="1,0.5"]
    
    Music.Lessons -> High.grades 
    Family.income -> Music.Lessons
    Family.income -> High.grades
}')
plot(g)
```
