---
title: 'L08: Samples and Observational studies'
output: pdf_document
---

<!-- libraries -->
```{r,include=FALSE,purl=FALSE}
library(knitr) # for include_graphics() 
library(dplyr)
```
### Observation vs. Experimentation
- Suppose we are interested in whether a certain surgical procedure prolongs life in cancer patients:
&nbsp;
```{r dag, echo=F, fig.height=1.5, fig.width=3,warning=FALSE}
#students, you don't need to know the dagitty package or understand this code
library(dagitty)

g <- dagitty('dag {
    Surgery [pos="0,1"]
    Death [pos="2,1"]
    Disease.severity [pos="1,0.5"]
    
    Surgery -> Death 
    Disease.severity -> Surgery
    Disease.severity -> Death
}')
plot(g)
```


### Example of SRS in R

- First read in the hospital cesarean data

```{r, message=F, echo=FALSE}
library(readxl)
library(dplyr)
CS_data <- read_xlsx("kozhimannil.xlsx", sheet = 1)
CS_data <- CS_data %>% mutate(ID = row_number())
head(CS_data)
```


### Example of SRS in R

```{r sample-n-100}
CS_100_1 <- CS_data %>% sample_n(100)
CS_100_2 <- CS_data %>% sample_n(100)
```

Do you expect `head(CS_100_1)` to equal `head(CS_100_2)`?

### Example of SRS in R

```{r are-first-rows-the-same}
head(CS_100_1 %>% select(Births, HOSP_BEDSIZE, cesarean_rate, ID))
```

### Example of SRS in R
```{r are-first-rows-the-same2}
head(CS_100_2 %>% select(Births, HOSP_BEDSIZE, cesarean_rate, ID))
```

### Example of SRS in R
```{r are-first-rows-the-same3}
identical(CS_100_1, CS_100_2)
```


### Example of SRS in R

What if you want to ensure that you pick the same SRS as a friend?

Then you need to use `set.seed()`:

```{r}
set.seed(123)
CS_100_1 <- CS_data %>% sample_n(100)

set.seed(123)
CS_100_2 <- CS_data %>% sample_n(100)

identical(CS_100_1, CS_100_2)
```

### SRS a fraction in R

```{r}
CS_5percent <- CS_data %>% sample_frac(0.05)
```

### Proportionate Stratified sampling in R

- Group *and then* sample the same fraction from each group

```{r}
CS_10percent_grouped <- CS_data %>% 
  group_by(HOSP_BEDSIZE) %>%
  sample_frac(0.1)
dim(CS_10percent_grouped)
```

- Proportionate stratified SRS assembles a sample that maintains the relative 
proportions of `HOSP_BEDSIZE` in the chosen sample compared to the population

### Proportionate Stratified sampling in R

How to check you really did sample 10% of each `HOSP_BEDSIZE` group?

First see how many hospitals fall into each category in the original data
```{r ,out.width="80%"}
CS_data %>%
  count(HOSP_BEDSIZE)

CS_data %>% group_by(HOSP_BEDSIZE) %>% tally()

CS_data %>% group_by(HOSP_BEDSIZE) %>% summarize(n=n())
```

### Proportionate Stratified sampling in R
then in the sample
```{r ,out.width="80%"}
CS_10percent_grouped %>%
  ungroup() %>%
  count(HOSP_BEDSIZE)

CS_10percent_grouped %>% tally()
CS_10percent_grouped %>% group_by(HOSP_BEDSIZE) %>% tally()
CS_10percent_grouped %>% summarize(n=n())
```

