---
title: "L18: Sampling Distributions and the Central Limit Theorem"
---

```{r load-libraries, include=FALSE, message=F, warning=FALSE}
library(ggplot2)
library(dplyr)
```

## Sampling distribution of the mean

### A known Normal population

Suppose $X \sim N(100, 15)$. This is the population we will sample from. The solid
line marks the population mean, and the dashed lines sit three standard deviations
out on each side.

```{r pop-normal, out.width="60%", fig.align='center', message=F, warning=F}
ggplot(data = data.frame(x = c(50, 150)), aes(x)) +
  stat_function(fun = dnorm, n = 101, args = list(mean = 100, sd = 15), col = "orange") +
  labs(y = "density") +
  theme_minimal(base_size = 15) +
  geom_vline(xintercept = 145, lty = 2) +
  geom_vline(xintercept = 55, lty = 2) +
  geom_vline(xintercept = 100) +
  ggtitle("Distribution of X in the population.\nN(100, 15), dashed lines at +/- 3SDs")
```

### The two formulas to remember

For a simple random sample of size $n$, the sampling distribution of the mean
$\bar{x}$ is centered at $\mu$ and has standard deviation (the **standard error**)
$\frac{\sigma}{\sqrt{n}}$.

```{r se-mean}
# Example: IQ scores are N(100, 15). Take samples of size n = 25.
mu    <- 100
sigma <- 15
n     <- 25

se_mean <- sigma / sqrt(n)   # standard error of the sample mean
c(mean = mu, standard_error = se_mean)
```

So $\bar{x} \sim N(100, 3)$. The sample means are much less variable than the
individual observations (3 versus 15).

### What happens to the standard error as n grows?

```{r se-vs-n}
# Increasing n by a factor of 100 divides the standard error by sqrt(100) = 10.
sigma / sqrt(25)      # n = 25
sigma / sqrt(2500)    # n = 2500, one hundred times larger
```

## Sampling distribution of a proportion

The mean of the sampling distribution of $\hat{p}$ is $p$, and its standard
deviation is $\sqrt{\frac{p(1-p)}{n}}$. The Normal approximation is trustworthy
when both $np$ and $n(1-p)$ are larger than 10.

```{r se-prop}
p <- 0.30
n <- 50

se_prop <- sqrt(p * (1 - p) / n)   # standard error of the sample proportion
success_failure <- c(np = n * p, n_1_minus_p = n * (1 - p))

c(mean = p, standard_error = se_prop)
success_failure   # both should be > 10 to use the Normal approximation
```

## The Central Limit Theorem in action

We now sample from a strongly skewed population and watch the sampling
distribution of the mean become Normal as $n$ grows.

### A strongly skewed population

```{r clt-population, out.width = "60%", fig.align='center'}
ggplot(data = data.frame(x = c(0, 5)), aes(x)) +
  stat_function(fun = dexp, n = 101, args = list(rate = 1), col = "orange") +
  theme_minimal(base_size = 15) +
  labs(title = "Distribution of a random variable") +
  geom_vline(aes(xintercept = 1), lty = 2) +
  geom_text(x = 1.5, y = 0.75, label = "population mean", check_overlap = T)
```

### Build the population and a helper function

```{r clt-setup}
set.seed(142)
pop_data <- rexp(n = 100000, rate = 1)
pop_data <- as.data.frame(pop_data)
names(pop_data)[1] <- "var_1"

calc_sample_mean <- function(df) {
  df %>%
    summarize(sample_mean = mean(var_1))
}
```

### Sampling distribution of the mean, n = 2

1) Take a sample of size 2. This is very small.
2) Calculate its mean.
3) Repeat 1,000 times.
4) Plot the 1,000 means as a histogram.

The black dashed line is the population mean, the red dotted line is the mean of
the sampled means.

```{r clt-n2, fig.align='center', out.width="60%", warning=F}
many.sample.stats.n2 <- replicate(1000, sample_n(pop_data, 2), simplify = F) %>%
  lapply(., calc_sample_mean) %>%
  bind_rows() %>%
  mutate(sample.id = 1:n())

ggplot(many.sample.stats.n2, aes(x = sample_mean)) +
  geom_histogram(binwidth = 0.25, col = "white") +
  ggtitle("Histogram of sample means where n=2") +
  geom_vline(aes(xintercept = 1), lty = 2) +
  geom_vline(aes(xintercept = many.sample.stats.n2 %>% summarise(mean(sample_mean)) %>% pull), lty = 3, col = "red") +
  theme_minimal(base_size = 15) +
  scale_x_continuous(limits = c(0, 6))
```

### Sampling distribution of the mean, n = 10

```{r clt-n10, fig.align='center', out.width="60%", warning=F}
set.seed(143)
many.sample.stats.n10 <- replicate(1000, sample_n(pop_data, 10), simplify = F) %>%
  lapply(., calc_sample_mean) %>%
  bind_rows() %>%
  mutate(sample.id = 1:n())

ggplot(many.sample.stats.n10, aes(x = sample_mean)) +
  geom_histogram(binwidth = 0.1, col = "white") +
  ggtitle("Histogram of sample means where n=10") +
  geom_vline(aes(xintercept = 1), lty = 2) +
  geom_vline(aes(xintercept = many.sample.stats.n10 %>% summarise(mean(sample_mean)) %>% pull), lty = 3, col = "red") +
  theme_minimal(base_size = 15) +
  scale_x_continuous(limits = c(0, 6))
```

### Sampling distribution of the mean, n = 30

```{r clt-n30, fig.align='center', out.width="60%", warning=F}
set.seed(144)
many.sample.stats.n30 <- replicate(1000, sample_n(pop_data, 30), simplify = F) %>%
  lapply(., calc_sample_mean) %>%
  bind_rows() %>%
  mutate(sample.id = 1:n())

ggplot(many.sample.stats.n30, aes(x = sample_mean)) +
  geom_histogram(binwidth = 0.1, col = "white") +
  ggtitle("Histogram of sample means where n=30") +
  geom_vline(aes(xintercept = 1), lty = 2) +
  geom_vline(aes(xintercept = many.sample.stats.n30 %>% summarise(mean(sample_mean)) %>% pull), lty = 3, col = "red") +
  theme_minimal(base_size = 15) +
  scale_x_continuous(limits = c(0, 6))
```

The population is strongly skewed, yet by n = 30 the distribution of the sample
means looks Normal, centered on the true mean of 1. That is the Central Limit
Theorem.

## Check your understanding

A population has $\mu = 40$ and $\sigma = 12$. You take a simple random sample of
size $n = 36$.

- What is the standard error of the sample mean?
- What distribution does $\bar{x}$ approximately follow?

```{r check-understanding}
mu <- 40; sigma <- 12; n <- 36
se <- sigma / sqrt(n)     # 12 / 6 = 2
se
# By the CLT, xbar is approximately N(40, 2).
```
