---
title: "Probability Sampling"
date: "September 16, 2026"
#output: slidy_presentation
output: pdf_document
editor_options: 
  chunk_output_type: inline
---

## Applied Example: Using a Sample to Estimate Low Birthweight in U.S. Territories

### Description of the Data

These data are births by place of occurrence for U.S. territories (American Samoa,
Guam, North Mariana Islands, Puerto Rico, and U.S. Virgin Islands) from the year 2015.

This is a subset of the data downloaded from the [National Bureau of Economic Research](http://www.nber.org/data/vital-statistics-natality-data.html).
You can find more information about the data set in the [User Guide](ftp://ftp.cdc.gov/pub/Health_Statistics/NCHS/Dataset_Documentation/DVS/natality/UserGuide2016.pdf).

### Data Dictionary

Here is the data dictionary for this dataset:

| Variable        | Description                                                  |
|:----------------|:-------------------------------------------------------------|
| babyID          | Unique identifier: row number                                |
| dbwt            | Birth weight in Grams: 227-8165 grams          | 
| combgest        | Combined gestation, in weeks: 17th to 47th week of gestation|              
| sex             | Assigned sex at birth: M (Male) or F (Female)	                       |
| dob_mm          | Birth month                                    | 
| cig_rec         | If the mother reports smoking in any of the three trimesters of pregnancy she is classified as a smoker: (Y) Yes, (N) No, or (U) Unknown|

### Import the Data into R

```{r read-in-data, message=F,warning=F}
library(tidyverse)
birth_data <- read_csv(file = "./L03_US-territories-births.csv")
birth_data <- birth_data%>% select(-...1)
```

```{r dimensions}
head(birth_data)
```

### Overview of the Applied Example

* Take a **simple random sample** from the **population** of births. We will use
this sample to *estimate* the **sample proportion** of babies with low birth weight in
the population. 

* To know how close our **estimate** is to the **true proportion**,
we will first calculate the **population proportion** of an infant being born less that 5 pounds 8 
ounces, or 2,500 grams, which is the traditional cutoff used to classify an 
infant as having low birth weight. 

* In real life settings, we would not know the **true** proportion, we would
only know the value we estimate from our sample. But here, we can investigate
how close the sample estimate is to the true parameter and see what we can do
to make our estimate even better. 

* We will see how well we can estimate the true probability based on random 
samples of varying sizes.

### Step 1: Add a Variable to the Dataset for Low Birth Weight (LBW)

```{r}
birth_data <- birth_data %>% mutate(lbw = dbwt < 2500) 
```

* What does the variable `lbw` store?
* `lbw` stores "logical" values, which means it is either equal to `TRUE` or `FALSE`
* Variables that store only two values are called **binary** variables. They are
most commonly stored as logical data (TRUE/FALSE), numeric (0/1) or 
categorical ("Yes"/"No").
* In the above code, `R` evaluates whether birth weight (`dbwt`) is less than 2,500
grams for each birth. If it is, then `lbw = TRUE` for that birth, and if not then 
`lbw = FALSE` for that birth.

### Step 2: Calculate the Proportion of Low Birth Weight Infants in Population Across All U.S. Territories

```{r ans3}
lbw_population <- birth_data %>% summarize(true_prob_lbw = mean(lbw))
lbw_population

mean(birth_data$lbw)
```

* Question: How did we take a mean of values equal to `TRUE` or `FALSE`? 
* Answer: `R` treats `TRUE` as equivalent to `1` and `FALSE` as equivalent to `0`.
The mean of a variable coded as 0 or 1 is the **proportion** (or **probability**) of individuals who
have low birth weight.
* Remember: We do not usually know the true value because we rarely have data on
every individual in a population

### Step 3: Take a Random Sample of Size n=10

```{r}
random_sample_n10 <- birth_data %>% 
  slice_sample(n = 10) %>%
  mutate(sample_size = n()) 
#the last line of code adds the sample size to every row of the new data frame
#we will want to reference this information later
```

### Step 4: Progressively Increase the Sample Size and Store those Samples

Sample the rows of data using the following sample sizes.
Assign each of your samples to a different R object.

    1. 10
    2. 50
    3. 100
    4. 200
    5. 500
    6. 1000
    7. 5000
    8. 10000
    9. `r nrow(birth_data)` (i.e, the entire target population)
    
### Step 4 Code

```{r}
random_sample_n50 <- birth_data %>%   slice_sample(n = 50) %>% mutate(sample_size = n())

random_sample_n100 <- birth_data %>% slice_sample(n = 100) %>% mutate(sample_size = n())

random_sample_n200 <- birth_data %>% slice_sample(n = 200) %>% mutate(sample_size = n())

random_sample_n500 <- birth_data %>% slice_sample(n = 500) %>% mutate(sample_size = n())

random_sample_n1000 <- birth_data %>% slice_sample(n = 1000) %>% mutate(sample_size = n())

random_sample_n5000 <- birth_data %>% slice_sample(n = 5000) %>% mutate(sample_size = n())

random_sample_n10000 <- birth_data %>% slice_sample(n = 10000) %>% mutate(sample_size = n())

whole_pop <- birth_data %>% slice_sample(n = nrow(birth_data)) %>% mutate(sample_size = n())
```

### Step 4 Side Note

By default `slice_sample(n=100)` takes a sample of size 100 **without replacement**.
This means that each individual can only be included in the sample at most one time.

In future classes, we will introduce methods where we select a sample 
**with replacement**. We will talk more about how this works in the coming weeks.

### Step 5: Calculate the Estimate of the Proportion of LBW for Each Random Sample

* For each sample, we want to estimate the proportion of `LBW` babies to see how
much it differs from the true proportion in the entire population
* We could do this using `summarize` for each sample and by writing that code 
ten times, but there is an easier way

### Step 5: Code to Estimate the Proportions More Efficiently

The function `bind_rows(df1, df2, df3, ...)` can be used to stack multiple data 
frames (e.g. df1, df2, df3, ...) on top of each other when they all contain 
the same variables. Bind together the nine data frames created in the previous code
chunk using `bind_rows()`and assign the stacked data frame the name `stacked_samples`:

```{r stacked-samples}
stacked_samples <- bind_rows(random_sample_n10, random_sample_n50, 
                             random_sample_n100, random_sample_n200, 
                             random_sample_n500, random_sample_n1000, 
                             random_sample_n5000, random_sample_n10000, 
                             whole_pop)
```

### Step 5: Code to Estimate the Proportions More Efficiently

Estimate the proportion of babies with low birth weight using each of 
your samples in `stacked_samples`. Hint: `group_by()` and `summarize()` will 
come in handy! Assign the output to a data frame called `sample_estimates`

```{r ans5}
sample_estimates <- stacked_samples %>% 
  group_by(sample_size) %>% 
  summarize(estimated_proportion_lbw = mean(lbw)) 

sample_estimates
```

### Step 6: Visualize the Results

* Make a line plot of the estimates of the proportions versus the sample size
* Add a horizontal line to the line plot at the **population proportion** that we are striving to
estimate
* We also add points on top of the line to see exactly where the estimates are

```{r ans6, out.width="60%"}
ggplot(sample_estimates, aes(x = sample_size, y = estimated_proportion_lbw)) + 
  geom_line(col = "blue") + 
  geom_point(col = "blue") +
  geom_hline(yintercept = lbw_population %>% pull(true_prob_lbw)) +
  labs(y = "Sample estimate of LBW", x = "Sample size") + 
  theme_minimal(base_size = 15) 
```

### Step 6: Visualize the Results

* Because the scale of the x axis is so large, try using `scale_x_log10()`
to convert the x-axis to a logarithmic scale

```{r}
ggplot(sample_estimates, aes(x = sample_size, y = estimated_proportion_lbw)) + 
  geom_line(col = "blue") + 
  geom_point(col = "blue") +
  geom_hline(yintercept = lbw_population %>% pull(true_prob_lbw)) +
  scale_x_log10() +
  labs(y = "Sample estimate of LBW", x = "Sample size") + 
  theme_minimal(base_size = 15) 
```

### Check Your Understanding!

1) What happens as the sample size increases? 

2) Will it **always** be the case that a higher sample size produces an estimate 
closer to the true value than a lower sample size?

### sampling variability

```{r}
nrep = 2000
estimates_n_100 = rep(NA,nrep)
for (i in 1:nrep){
  set.seed(i)
  random_sample_n100 <- birth_data %>% slice_sample(n = 100)
  estimates_n_100[i] <- mean(random_sample_n100$lbw)
}

histogram_100 <- ggplot(data = data.frame(estimates = estimates_n_100), aes(x=estimates)) + geom_histogram()
histogram_100

mean(estimates_n_100)
sd(estimates_n_100)

mean(estimates_n_100)
mean(birth_data$lbw)
```

```{r}
nrep = 2000
estimates_n_500 = rep(NA,nrep)
for (i in 1:nrep){
  set.seed(i)
  random_sample_n500 <- birth_data %>% slice_sample(n = 500)
  estimates_n_500[i] <- mean(random_sample_n500$lbw)
}

histogram_500 <- ggplot(data = data.frame(estimates = estimates_n_500), aes(x=estimates)) + geom_histogram()
histogram_500

mean(estimates_n_500)
sd(estimates_n_500)

mean(estimates_n_500)
mean(birth_data$lbw)
```

```{r}
nrep = 2000
estimates_n_2000 = rep(NA,nrep)
for (i in 1:nrep){
  set.seed(i)
  random_sample_n2000 <- birth_data %>% slice_sample(n = 2000)
  estimates_n_2000[i] <- mean(random_sample_n2000$lbw)
}

histogram_2000 <- ggplot(data = data.frame(estimates = estimates_n_2000), aes(x=estimates)) + geom_histogram()
histogram_2000

mean(estimates_n_2000)
sd(estimates_n_2000)

mean(estimates_n_2000)
mean(birth_data$lbw)
```

```{r}

all_estimates = data.frame(estimates = c(estimates_n_100,estimates_n_500,estimates_n_2000), sample_size = rep(c(100,500,2000), each=2000))
ggplot(data = all_estimates,aes(x=estimates, fill=as.factor(sample_size))) + 
  geom_histogram(alpha=0.5, position="identity")

all_estimates %>% group_by(sample_size) %>% summarize(mean_estimate=mean(estimates), sd_estimate=sd(estimates))
```